---
title: "Project 1"
author: "Michael Wiley mw3239"
format:
  revealjs:
    incremental: false
    scrollable: true
    theme: dark
    code-fold: true
    css: styles.css
  html:
    code-fold: true
---

# Happy Moments {.center}

A Lingustic Deep Dive

## Imports and Setup

-   Relevant `tidyverse` libraries
-   `here` used for relative directory navigation
-   `emojifont` used to place emoji on plots.
-   `plotly` and `shiny` used to create interactive visuals.

```{r imports, message = FALSE}
#| echo: true

library(magrittr)
library(stringr)
library(readr)
library(dplyr)
library(ggplot2)

library(here)
library(emojifont)
library(plotly)
library(shiny)
```

## Load data {visibility="hidden"}

Loads helper functions from the `lib` folder.

```{r source files, echo = FALSE, warning = FALSE, message = FALSE}
#| echo: true
#| message: false
#| warning: false

source_files <- list.files(here("lib")) %>%
  str_subset("\\.R$") %>%
  str_c(here("lib"), "/", .)

lapply(source_files, source)
```

```{r load data, warning = FALSE, message = FALSE}
#| echo: true

load_data()
```

# Question 1:

Within the happy moments dataset, do moments belonging to different categories feature different part of speech distributions? Perhaps some categories feature an uncharacteristic number of adjectives, for example?

## Filter data

Keep only categories we know the ground truth for - there's still a lot of data and it's not clear how accurate the predicted categories are.

```{r q1_filter}
#| echo: true

filtered_data <- cleaned_hm %>%
  left_join(senselabel, by = "hmid") %>%
  filter(!is.na(ground_truth_category))
```

## Process data

Calculate relative frequency of POS within each ground_truth_category. This is necessary since the ground_truth_category counts are not at all balanced.

```{r q1_process, message = FALSE, warning = FALSE}
#| echo: true
#| message: false

pos_count <- filtered_data %>%
  group_by(ground_truth_category, POS) %>%
  summarise(count = n()) %>%
  group_by(ground_truth_category) %>%
  mutate(relative_frequency = count / sum(count)) %>%
  arrange(ground_truth_category, POS)
```

## Check data

```{r q1_check}
#| echo: true

pos_count
```

## Write data {visibility="hidden"}

```{r q1_write, message = FALSE}
write_csv(pos_count, here("output", "pos_count.csv"))
```

## Visualize data

Plot the faceted bar chart of relative frequency for each POS and ground_truth_category

```{r q1_plot, message = FALSE}
#| echo: true

ggplot(pos_count, aes(x = ground_truth_category, y = relative_frequency, fill = ground_truth_category)) +
  geom_bar(stat = "identity") +
  xlab("Ground Truth Category") +
  ylab("Relative Frequency") +
  ylim(0, max(pos_count$relative_frequency)) +
  ggtitle("Relative Frequency of Ground Truth Categories by POS") +
  facet_wrap(~ POS, nrow = 2) +
  scale_fill_discrete(name = "Ground Truth Category") +
  dark_theme() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
        axis.text.y = element_text(size = 12),
        axis.title.x = element_text(size = 17),
        axis.title.y = element_text(size = 17)
        )

ggsave(here("figs", "categories_by_pos.png"))
```

::: aside
:::

## Question 1 Analysis:

While there isn't quite as much variance as I'd have hoped, the bar charts still show some that every category uses each part of speech in quite the same way.

The big standouts to me here are affection pronounces and leisure nouns, but lets dive a little deeper.

# Question 2:

Which (POS, ground_truth_category) pairs most differ from other categories for that part of speech?

Is there a category that frequently differs from other categories across many parts of speech?

## Process data

-   Calculate, on average, what percentage a given POS occurs for each category.
-   Calculate difference from average for each POS, prediction pair

```{r q2_process}
#| echo: true

average_pos_frequency <- pos_count %>%
  group_by(POS) %>%
  summarise(average_frequency = mean(relative_frequency))

average_pos_frequency <- pos_count %>%
  left_join(average_pos_frequency, by = "POS") %>%
  mutate(difference_from_average = relative_frequency - average_frequency)
```

## Add emoji

```{r q2_process_emoji}
#| echo: true

average_pos_frequency <- average_pos_frequency %>%
  mutate(emoji = case_when(ground_truth_category=="achievement"~"🏆",
                           ground_truth_category=="affection"~"❤️",
                           ground_truth_category=="bonding"~"👪",
                           ground_truth_category=="enjoy_the_moment"~"😊",
                           ground_truth_category=="leisure"~"⛱️",
                           ground_truth_category=="nature"~"🌳",
                           ground_truth_category=="exercise"~"🏋️",
                           T~"")
         )
```

## Write data

```{r q2_write}
#| echo: true

write_csv(average_pos_frequency, here("output", "average_pos_frequency.csv"))
```

## Check data

```{r print_test}
#| echo: true

average_pos_frequency
```

## Plot data

```{r q2_plot}
#| echo: true

p <- ggplot(average_pos_frequency, aes(x = POS, y = difference_from_average, label = emoji, color = ground_truth_category, text = ground_truth_category)) +
  geom_text(family = "EmojiOne", size = 7) +
  labs(title = "Difference in Relative Frequency from Average",
       x = "POS", y = "Difference") +
  dark_theme() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  guides(color = FALSE)

p <- ggplotly(p, tooltip = c("x", "y", "text"))

p
```

```{r q2_save}
html_file <- here("figs", "relative_freq_vs_avg.html")
htmlwidgets::saveWidget(p, file = html_file, selfcontained = TRUE)
```

::: aside
The chart is interactable! Hover over an emoji to see which category it belongs to.
:::

## Question 2 Analysis:

This chart confirms what we initially observed from the bar chart from the initial research question.

Pronouns and nouns appears to be the parts of speech most affected by the category, while adjectives and conjunctions are hardly affected at all.

## Standouts

Occur More than in Other Categories:

-   Affection + Pronouns
-   Leisure + Nouns
-   Bonding + Nouns

Occur Less than in Other Categories:

-   Nature + Pronouns
-   Exercise + Nouns
-   Affection + Determiners

# Questions 3 and 4:

-   What words occur most frequently in these outlier (POS, category) pairs?
-   Likewise, what words occur more frequently in a given (POS, category) than in other categories?

## Filter data

-   Join `senselabel` with `cleaned_hm` on `hmid` and once again calculate frequencies
-   Calculate average relative frequency across all categories for each `POS`
-   Finally add an average relative frequency column to the `joined_data` df

```{r q3_filter, message = FALSE}
#| echo: true

joined_data <- inner_join(cleaned_hm, senselabel, by = "hmid") %>%
  dplyr::filter(!is.na(ground_truth_category)) %>%
  group_by(POS, lowercaseLemma) %>%
  mutate(total_count = n()) %>%
  ungroup() %>%
  group_by(POS, lowercaseLemma, ground_truth_category) %>%
  mutate(within_category_count = n()) %>%
  ungroup() %>%
  mutate(relative_frequency = within_category_count / total_count)


average_relative_freq <- joined_data %>%
  group_by(POS, lowercaseLemma) %>%
  summarize(avg_relative_freq = mean(relative_frequency, na.rm = TRUE))


joined_data <- joined_data %>%
  left_join(average_relative_freq, by = c("POS", "lowercaseLemma")) %>%
  group_by(POS, lowercaseLemma) %>%
  mutate(max_relative_frequency = max(relative_frequency))
```

## Write data

```{r q3_write}
write_csv(joined_data, here("output", "word_freq_analysis.csv"))
```

## Set Top Word Default {visibility="hidden"}

```{r q3_constant}
top_words_count <- 10
```

## Shiny

Sets up the shiny ui and server.

```{r q3_shiny_ui}
#| echo: true

ui <- fluidPage(
  titlePanel("Word Frequency Analysis"),
  
  sidebarLayout(
    sidebarPanel(
      # Category filter
      selectInput("category_filter", "Category:", choices = unique(joined_data$ground_truth_category)),
      
      # POS filter
      selectInput("pos_filter", "POS:", choices = unique(joined_data$POS)),
      
      # Top words count
      numericInput("top_words_count", "Top Words Count:", value = 10, min = 1),
      
      # Standout words threshold
      sliderInput("standout_words_threshold", "Standout Words Threshold:", min = 0, max = 1, value = 0.8),
      
      #Total Count Filter      
      sliderInput("total_count_filter", "Total Count Filter:", min = 1, max = 100, value = 2)
    ),
    
    mainPanel(
      tabsetPanel(
        tabPanel(
          "Top Words",
          tableOutput("top_words_table")
        ),
        
        tabPanel(
          "Standout Words",
          p("This tab displays the words that appear in the selected category significantly more than any other category. The relative_frequency denotes how many times a word appeared for a particular (POS, category) combination relative to every other category. For example, a relative frequency of 0.8 and total_count of 5 means of the 5 times the word appeared, 4 of them were in this category."),
          tableOutput("standout_words_table")
        )
      )
    )
  )
)
```

```{r q3_shiny_server}
#| echo: true

server <- function(input, output) {
  output$top_words_table <- renderTable({
    filtered_data <- joined_data %>%
      filter(POS == input$pos_filter, ground_truth_category == input$category_filter) %>%
      count(lowercaseLemma) %>%
      arrange(desc(n)) %>%
      head(input$top_words_count)
    filtered_data
  })
  
  output$standout_words_table <- renderTable({
    filtered_data <- joined_data %>%
      dplyr::filter(POS == input$pos_filter,
             ground_truth_category == input$category_filter,
             total_count >= input$total_count_filter) %>%
      group_by(lowercaseLemma) %>%
      dplyr::select(lowercaseLemma, relative_frequency, avg_relative_freq, total_count) %>%
      distinct() %>%
      dplyr::filter(relative_frequency >= input$standout_words_threshold) %>%
      arrange(desc(relative_frequency), desc(total_count))
    filtered_data
  })
}
```

## App

```{r q3_run_shiny}
#| echo: true
#| context: server

shinyApp(ui = ui, server = server)
```

::: aside
Note: This must be run directly from `Project1.qmd`
:::

## Questions 3 & 4 Analysis

-   `I` was the most common pronoun for all happy moment categories... besides `affection`, where `my` came out on top.
-   Surprisingly, however, not too many words stood out to me as appearing with significantly high counts only in specific categories that I wouldn't expect to be there. What stood out to me more is that a number of words seem to have the wrong part of speech attached to them entirely.
-   Try filtering `POS` down to `ADJ` on the `Standout Words` tab and observe how many non-adjectives were classified as adjectives.

# Conclusion and Future Analysis

-   This analysis focused on how sentence construction and POS distribution varied within the context of happy moments. However, it might be interesting to perform this same analysis at a higher level to see how the composition of happy moments sentences as a whole compare to sentences in the English language in general.
